---
title: "Untitled"
author: "Bowie Lam"
date: "12/26/2020"
output:
  html_document:
    df_print: paged
  pdf_document:
    keep_html: yes
---

```{r}
# load packages
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
library(readr)
library(stringr)
library(here)
```

```{r}
# import data
fires <- read.csv(here("raw_data", "mapdataall.csv"), na = "", stringsAsFactors = FALSE)
```

```{r}
# view the structure of the data set
str(fires)
```

```{r}
# view a summary of the data set
summary(fires)
```

```{r}
# remove any duplicated rows
fires <- fires[!duplicated(fires), ]
```

```{r}
# display the proportion of missing data for each column
pctmiss <- colSums(is.na(fires))/nrow(fires)
round(pctmiss, 2)
```

```{r}
# remove columns with a high proportion of missing data
fires$incident_control <- NULL
fires$incident_administrative_unit_url <- NULL
fires$incident_type <- NULL

# remove columns that are not needed/necessary for the purposes of this analysis
fires$incident_location <- NULL
fires$incident_id <- NULL
fires$incident_date_extinguished <- NULL
fires$notification_desired <- NULL
fires$incident_is_final <- NULL
fires$incident_url <- NULL
fires$incident_date_last_update <- NULL
fires$incident_date_created <- NULL
fires$is_active <- NULL
fires$calfire_incident <- NULL
```

### Inspecting Each Variable

> In this section, each variable, or column, is inspected for any errors, anomalies, or other details worth knowing before producing a cleaned data set for analysis.

#### variable(s): `incident_name`

```{r}
# trim any white space
fires$incident_name <- trimws(fires$incident_name)

# number of rows in the data set
nrow(fires) # 1732

# number of uniquely named rows in the data set
fires %>% pull(incident_name) %>% unique() %>% length() # 1264
```

The data set has a total of 1,732 rows, but it has 1,264 uniquely named rows, thus indicating there are incidents with the same name, either because (1) they are the same incidents that occurred at different times, (2) unrelated incidents received the same name, or (3) some other explanation. This requires further exploration into the observations that share the same names.

#### variable(s): `incident_dateonly_extinguished` and `incident_dateonly_created`

```{r}
# note: there is >= 1 observation that does not have data in the `incident_dateonly_extinguished` column
# they're denoted by `NA`
nrow(fires %>% filter(is.na(incident_dateonly_extinguished))) > 0
```

```{r}
# note: there are no observations that do not have data in the `incident_dateonly_created` column; in other words, all observations have data in the `incident_dateonly_created` column
nrow(fires %>% filter(is.na(incident_dateonly_created))) > 0
```

```{r}
# separate the date data in `incident_dateonly_extinguished` into three individual columns
fires <- separate(fires, "incident_dateonly_extinguished", c("extinguished_year", "extinguished_month", "extinguished_day"), sep = "-")

# separate the date data in `incident_dateonly_created` into three individual columns
fires <- separate(fires, "incident_dateonly_created", c("created_year", "created_month", "created_day"), sep = "-")
```

```{r}
# remove leading zero (0) and convert from chr to num for columns `extinguished_month`, `extinguished_day`, `created_month`, and `created_day`
fires$extinguished_month <- as.numeric(str_extract(fires$extinguished_month, "(?!0)[0-9]+"))
fires$extinguished_day <- as.numeric(str_extract(fires$extinguished_day, "(?!0)[0-9]+"))
fires$created_month <- as.numeric(str_extract(fires$created_month, "(?!0)[0-9]+"))
fires$created_day <- as.numeric(str_extract(fires$created_day, "(?!0)[0-9]+"))

# convert `extinguished_year` and `created_year` from chr to num
fires$extinguished_year <- as.numeric(fires$extinguished_year)
fires$created_year <- as.numeric(fires$created_year)
```

#### variable(s): `incident_administrative_unit`

```{r}
# trim any white space
fires$incident_administrative_unit <- trimws(fires$incident_administrative_unit)
```

#### variable(s): `incident_county`

```{r}
# trim any white space
fires$incident_county <- trimws(fires$incident_county)
```

#### variable(s): `incident_cooperating_agencies`

```{r}
# trim any white space
fires$incident_cooperating_agencies <- trimws(fires$incident_cooperating_agencies)
```

```{r}
str(fires)
```

```{r}
fires
```





