---
title: "Untitled"
author: "Bowie Lam"
date: "12/26/2020"
output:
  html_document:
    df_print: paged
  pdf_document:
    keep_html: yes
---

```{r}
# load packages
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
library(readr)
library(stringr)
library(here)
```

```{r}
# import data
fires <- read.csv(here("raw_data", "mapdataall.csv"), na = "", stringsAsFactors = FALSE)
```

```{r}
# view the structure of the data set
str(fires)
```

```{r}
# view a summary of the data set
summary(fires)
```

```{r}
# remove any duplicated rows
fires <- fires[!duplicated(fires), ]
```

```{r}
# display the proportion of missing data for each column
pctmiss <- colSums(is.na(fires))/nrow(fires)
round(pctmiss, 2)
```

```{r}
# remove columns with a high proportion of missing data
fires$incident_control <- NULL
fires$incident_administrative_unit_url <- NULL
fires$incident_type <- NULL

# remove columns that are not needed/necessary for the purposes of this analysis
fires$incident_location <- NULL
fires$incident_id <- NULL
fires$incident_date_extinguished <- NULL
fires$notification_desired <- NULL
fires$incident_is_final <- NULL
fires$incident_url <- NULL
fires$incident_date_last_update <- NULL
fires$incident_date_created <- NULL
fires$is_active <- NULL
fires$calfire_incident <- NULL
fires$incident_administrative_unit <- NULL
fires$incident_cooperating_agencies <- NULL
```

### Inspecting Each Variable

> In this section, each variable, or column, is inspected for any errors, anomalies, or other details worth knowing before producing a cleaned data set for analysis.

#### variable(s): `incident_name`

```{r}
# trim any white space
fires$incident_name <- trimws(fires$incident_name)

# number of rows in the data set
nrow(fires) # 1732

# number of uniquely named rows in the data set
fires %>% pull(incident_name) %>% unique() %>% length() # 1264
```

The data set has a total of 1,732 rows, but it has 1,264 uniquely named rows, thus indicating there are incidents with the same name, either because (1) they are the same incidents that occurred at different times, (2) unrelated incidents received the same name, or (3) some other explanation. This requires further exploration into the observations that share the same names.

```{r}
# find the names of fires that occur more than once in the data set
repeated_names <- fires %>% group_by(incident_name) %>% count() %>% arrange(desc(n)) %>% filter(n > 1) %>% pull(incident_name)

# produce a data frame where `incident_name` matches all of those in `repeated_names`
# since duplicated rows where already removed, this step is meant to discover why there are repeated names
fires %>% filter(incident_name %in% repeated_names) %>% arrange(incident_name)
```

After analyzing the data for many observations that share the name `incident_name`, it appears that the reason for repeated names is simply due to the fact that the name was used more than once for an unrelated incident. Therefore, as far as I can tell at this point, no manipulations need to be made to this section.

#### variable(s): `incident_county`

```{r}
# trim any white space
fires$incident_county <- trimws(fires$incident_county)
```

```{r}
fires %>% group_by(incident_county) %>% count() %>% arrange(desc(n))
```

#### variable(s): `incident_acres_burned`

```{r}
fires %>% group_by(incident_acres_burned) %>% count() %>% filter(incident_acres_burned < 1 | is.na(incident_acres_burned))
```

Acres burned is either zero (`0`) or `NA` for only a small proportion of the data set, which is insigificant in the grand scheme of things. Therefore, this variable does not require any manipulation, and will remain in the data frame, just in case.

#### variable(s): `incident_containment`

```{r}
fires %>% group_by(incident_containment) %>% count() %>% arrange((desc(incident_containment)))
```

For the most part, incidents are nearly 100% contained. Only 22 observations do not have this information, but they're very insignificant when compared to those with this information available. Therefore, this variable does not require any manipulation, and will remain in the data frame, just in case.

#### variable(s): `incident_longitude` and `incident_latitude`

```{r}
# find the value(s) of the observations whose longitude does not match the longitude pattern
matchless_lons <- fires$incident_longitude[str_detect(fires$incident_longitude, "[-]?[0-9]{2,}\\.?[0-9]*", 
                    negate = TRUE)]

# find the value(s) of the observations whose latitude does not match the longitude pattern
matchless_lats <- fires$incident_latitude[str_detect(fires$incident_latitude, "[-]?[0-9]{2,}\\.?[0-9]*", 
                    negate = TRUE)]
```

```{r}
# using the previous step, find the observation(s) whose longitude and/or latitude do not match the pattern
matchless_fires <- fires %>% filter(incident_longitude %in% matchless_lons | incident_latitude %in% matchless_lats)
matchless_fires
```

Out of the entire data set, the Pala Fire is the only observation whose longitude and/or latitude do not match the typical pattern. Since this is the only one, I decided to manually input the correct coordinates for this observation. 

```{r}
# since there's only one matchless fire, reassign it to the specific fire in question
pala_fire <- matchless_fires

# insert longitude and latitude of the Pala Fire
pala_fire$incident_longitude <- -117.11794
pala_fire$incident_latitude <- 33.36109

# remove the old Pala Fire data from the data frame
fires <- fires[!fires$incident_name == "Pala Fire",]

# insert the new Pala Fire data into the data frame
fires <- rbind(fires, pala_fire)

# make sure the main data frame has the new Pala Fire data
fires %>% filter(incident_name == "Pala Fire")
```

#### variable(s): `incident_dateonly_extinguished` and `incident_dateonly_created`

The data is currently in YYYY-MM_DD, which can be difficult for analysis. Therefore, the data for both columns will be separated into individual columns that contain year, month, and day.

```{r}
# note: there is >= 1 observation that does not have data in the `incident_dateonly_extinguished` column
# they're denoted by `NA`
nrow(fires %>% filter(is.na(incident_dateonly_extinguished))) > 0
```

```{r}
# note: there are no observations that do not have data in the `incident_dateonly_created` column; in other words, all observations have data in the `incident_dateonly_created` column
nrow(fires %>% filter(is.na(incident_dateonly_created))) > 0
```

```{r}
# separate the date data in `incident_dateonly_extinguished` into three individual columns
fires <- separate(fires, "incident_dateonly_extinguished", c("extinguished_year", "extinguished_month", "extinguished_day"), sep = "-")

# separate the date data in `incident_dateonly_created` into three individual columns
fires <- separate(fires, "incident_dateonly_created", c("created_year", "created_month", "created_day"), sep = "-")
```

```{r}
# remove leading zero (0) and convert from chr to num for columns `extinguished_month`, `extinguished_day`, `created_month`, and `created_day`
fires$extinguished_month <- as.numeric(str_extract(fires$extinguished_month, "(?!0)[0-9]+"))
fires$extinguished_day <- as.numeric(str_extract(fires$extinguished_day, "(?!0)[0-9]+"))
fires$created_month <- as.numeric(str_extract(fires$created_month, "(?!0)[0-9]+"))
fires$created_day <- as.numeric(str_extract(fires$created_day, "(?!0)[0-9]+"))

# convert `extinguished_year` and `created_year` from chr to num
fires$extinguished_year <- as.numeric(fires$extinguished_year)
fires$created_year <- as.numeric(fires$created_year)
```

```{r}
str(fires)
```







